A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-31 latex/main.tex
Updates from Overleaf

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, using small numbers of widely diverged
genomes. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of genomes sampled from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to perform species-level tree
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically built with
simpler distance methods [citation?]
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations,
but also because of the complexity of modeling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Guo2022recombination,
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support that explicitly \emph{uses}
recombination-aware inference is mostly limited to
visualization, for example of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
or tree changes along the genome~\cite{Rasmussen2023}.

@@

The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
corresponds to nodes that are not associated with a recombination event.
These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence
to suggest recombination (Document \ref{sec:dd_xn}).
XB, on the other hand,
has been thoroughly characterized \cite{Gutierrez2022} and
may have misclassified by sc2ts due to limitations in
handling deletions or matching against long branches (Document \ref{sec:dd_xb}). 
See Document~\ref{sec:supplement_indepth_pango_x_lineages}
for extended analyses on all of the Pango X lineages,
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns
associated with Table~\ref{tab:pango_x_lineages}, and
Document S2 for ARG visualizations of the Pango X lineages.
[TODO ref Document \ref{sec:supplement_rejected_pango_x_lineages}
to mention multi breakpoints and close issue 379]

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG
(shading in Figure~\ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group
of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants
inferred within each week along the ARG,
focusing on the 354 events passing QC filters.
Recombination events were inferred more often
during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure~\ref{fig:recombinants_over_time}B; 
% JK is this correct?? we had 1.0 + 2.7 \; 10^{-7} x$ which I found confusing
linear regression: $1.0 + 2.7\times 10^{-7} x$, $p = 1.1 \times10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure~\ref{fig:recombinants_over_time}C; linear regression: 
$0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

@@

% P1. The ARG. Unique resource. Emphasise utility of UShER and how 
% this has more.

We present the first pandemic-scale ancestral recombination graph (ARG) for
SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
ARG represents a step change in how the evolutionary history of this virus can
be represented and analysed. Like the UShER phylogeny, it provides a
comprehensive and continuously updatable record of global genomic diversity,
but it extends this paradigm by embedding recombination directly into the
genealogy. The ARG unifies point mutations, deletions, and recombination events
within a single, compact data structure, enabling analyses that
were previously fragmented across multiple methods and data models. By encoding
reticulate ancestry explicitly rather than assuming a single tree, the ARG
provides a complete and internally consistent representation of SARS-CoV-2
evolution. In this way, the
ARG offers the same accessibility and scalability that made the UShER tree
indispensable during the pandemic, while expanding its scope to encompass the
full spectrum of evolutionary processes shaping SARS-CoV-2.

% P2. This is a conservative lower-bound on recombination. 
% Discuss other estimates.
% 
% SHZ: Why should we care about getting this number/frequency right?
% Sure, popgen folks love estimating this, but to other people,
% existing estimates already tell us that
% recombination in SARS-CoV-2 is (qualitatively) rare,
% even without the advances of sc2ts.
% Also, we are talking about the frequency of detectable recombination
% in the last sentence, right?
% The true frequencey of recombination is unknowable from this type of data.
% So, the best we are ever going to get is a better lower bound.
The recombination frequency estimated from the \texttt{sc2ts} ARG is
substantially lower than that implied by the combination of UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
reported 589 unique recombination events with 43,104 descendant
samples---corresponding to roughly 2.7\% of all genomes being of recombinant
origin.  Using the same time interval, our inference at \(k=4\) identifies only
14 high-confidence recombination events with 172 descendants among 501,235
genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
explained by the stricter parsimony penalty adopted here: increasing the
penalty from \(p=3\) (as used in the RIPPLES analysis) 
to \(k=4\) greatly reduces false
positives while retaining nearly all verified recombinant lineages, 
including those missed by RIPPLES at equivalent thresholds.  
We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
of recombination as
tree rearrangements proceed, thereby requiring the more lenient 
threshold in order to find known recombinats.
These comparisons suggest that our estimate represents a conservative 
lower bound. Further work will be required to determine
the true frequency of recombination in SARS-CoV-2 evolution.

% P3. Clarifies relationships between pango X lineages, simplifying
% the Omicron sublineage.
% 
% SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
% Sure, phylogeneticists would get a kick out of it,
% but I don't think the majority of Cell readership would.
% I think what is important is knowing
% how many emerging recombinants carrying new combinations of mutations of concern
% we are actually dealing with.
% Clarifying the relationships between Pango Xs make this possible.

@@

% P5. Confirm breakpoints around Spike. 
% SHZ: Maybe we can say it's a known (instead of robust)
% pattern of detectable recombination in CoVs in general?
% There are some citations for template switching hotspots in SC2.
% Also, it's not only about selection favoring new combinations of mutations in Spike.
% It's about combinations of mutations in, say, ORF1ab, getting selected against.
% Maybe mention epistasis?
[FIXME: This is directish output from ChatGPT. Review!]
The genomic distribution of breakpoints in the ARG is highly uneven, with a
pronounced enrichment near the $5'$ boundary of the Spike gene and across the
S1–S2 junction.  This pattern closely mirrors that observed in the
UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
analyses \citep{Jackson2021} [more cites], but here it is recovered
independently over a much longer sampling period and under a more conservative
inference regime.  Under our more conservative estimate of the overall
prevalence of recombination, we find that this enrichment persists despite a substantial
reduction in the total number of inferred recombination events, indicating that
it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
by-product of detection sensitivity.  Across all high-confidence events, the
median breakpoint interval length is $\sim 2.2$kb and declines with parental
divergence, consistent with expectations from sequence-based localization
accuracy.  These results show that even with a conservative lower bound on
recombination frequency, the concentration of breakpoints around Spike remains
a robust signature of coronavirus genome biology—likely reflecting a
combination of mechanistic biases in template switching and selective
constraints on Spike function.

@@

@@

[TODO]

We close by emphasizing several current limitations and potential directions to improve sc2ts.
First, more stringent filters to exclude false positive recombinants during inference
would obviate post-hoc treatment of these recombinants.
Second, indels carry important phylogenetic signals,
for example, as defining mutations of Alpha and Delta (REF).
We currently treat multi-base deletions as missing data or multiple independent deletions,
but they are in reality single events.
More sophisticated treatments of multi-base indels in the HMM could help recover additional signals
(and avoid false signals) of recombination.
Third, the current HMM assumes a simplistic mutation model,
which does not account for documented transition-to-transversion rate bias (REF)
and context-dependent variation in mutation rate (REF).
A more realistic mutation model may help improve inference of local genetic relationships in the ARG.
Fourth, an inherent limitation of detecting recombination from sequence data,
which applies to all methods, is that it is impossible to detect events with few (or no) supporting bases,
as these cannot be distinguished from mutation events.
This limitation makes it particularly challenging to detect events near the termini of linear viral genomes.
Fifth, it is challenging to accurately reconstruct global events
when there are regions with poor sampling and jurisdictions with long delays
before sequences become publicly available (REF).
Revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating with samples that became available later)
and better handling of putative recombinants with suspiciously long branches
(e.g., Delta and Omicron BA.2) might improve the accuracy of the genealogy.
Finally, future developments may account for geography in the HMM,
favoring attachment of samples to geographically close lineages.

TODO: add a short discussion of the node dates and some references
to the literature, closing issue 725

@@

We validated recombination events in the sc2ts ARG 
using 3SEQ~\cite{Lam2018improved} as follows.
We first exported the alignments for the recombination node and both 
parents to FASTA format, and then ran 3SEQ in ``full mode''
independently for each recombinant. Specifically, we 
ran \texttt{3seq -full  parents.fa recombinant.fa} using default 
thresholds.
We used 3SEQ source code downloaded from \url{https://gitlab.com/lamhm/3seq}
at git hash e45ee3b4.

@@

Recombination breakpoint intervals are imprecise
when the recombining parents are identical in sequence
at sites spanning the breakpoint.
Here, we estimate the interval lengths expected around a breakpoint,
given a genome-wide nucleotide substitution rate $\mu$,
total time between two samples ($T$), and
sufficient support to detect the recombination event.
If a breakpoint occurs a fraction $P$ along a genome of length $L$ and
$k$ mutations have occurred to the left of the breakpoint,
then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left.
If we require $n$ mutations on both sides of the breakpoint,
we can calculate the expected value of the interval to the left
by summing across a Poisson-distributed number of mutations,
conditioned on having at least $n$ mutations:

@@

To explore the timing of emergence of recombinants,
case count data were downloaded from
the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE)
at Johns Hopkins University \cite{Dong2020} and binned by week from January, 2020 to March, 2023.
The proportion of each lineage $i$ in each week ($p_i$) was then inferred
from the Scorpio designation at each node within the ARG for that week.
Genetic variation was measured as the ``expected heterozygosity'' across lineages,
which is the chance that two lineages chosen at random
from that week had different Scorpio designations ($\sum_{i} \sum_{j>i} 2 p_i p_j$).
The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time} allowed for a non-zero intercept,
because there may be unreported cases even without any reported cases.
The slopes were highly significant, and
the p-values were confirmed by randomly permuting the predictor variable 10,000 times.
ANOVAs were conducted to determine which single predictor ($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
These statistical tests and graphics were conducted
in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.

In addition to the analysis of the 354 recombination events 
that pass QC,
we analyzed the relationship between the full set of 855 recombinants and case numbers.
These additional recombinants occurred disproportionately during the Delta and first Omicron wave (July, 2021 to February, 2022).
Again, the number of recombinant events rose with the global number of cases
(linear regression: $3.1 + 5.3 \; 10^{-7} x$, $p = 0.0006$)
and with the diversity of lineages in each week
(linear regression: $0.42 + 18.3 x$, $p = 2 \; 10^{-12}$).
The single best predictor of recombination events among the predictors tested ($n$, $H$, $n \; H$, $n^2$, $n^2 H$)
was again the product of cases and diversity ($n \; H$),
with an adjusted $R^2$ of 26.9\%,
although this was now followed closely by expected heterozygosity on its own ($H$),
with an adjusted $R^2$ of 26.5\%.
Thus, similar conclusions are reached with either dataset,
although filtering the recombination events to the 354 with better support revealed a stronger relationship with case numbers.

@@

\subsection{Pango X lineages not present in the ARG}
\label{sec:supplement_rejected_pango_x_lineages}

>>>>

\section*{INTRODUCTION}
% SHZ: The first paragraph should explain
% why recombination "can" be important in pathogens.
% Then, we explain why we need recombination-aware methods for pathogens.
% This should set the stage for why this work is really important,
% not just for SC2 but more broadly for the field of pathogen genomics.

% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, using small numbers of widely diverged
genomes. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of genomes sampled from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to perform species-level tree
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically built with
simpler distance methods [citation?]
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations,
but also because of the complexity of modeling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Guo2022recombination,
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support that explicitly \emph{uses}
recombination-aware inference is mostly limited to
visualization, for example of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
or tree changes along the genome~\cite{Rasmussen2023}.

@@

The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
corresponds to nodes that are not associated with a recombination event.
These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence
to suggest recombination (Document \ref{sec:dd_xn}).
XB, on the other hand,
has been thoroughly characterized \cite{Gutierrez2022} and
may have misclassified by sc2ts due to limitations in
handling deletions or matching against long branches (Document \ref{sec:dd_xb}). 
See Document~\ref{sec:supplement_indepth_pango_x_lineages}
for extended analyses on all the Pango X lineages;
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns
associated with Table~\ref{tab:pango_x_lineages};
Document S2 for ARG visualizations of the Pango X lineages; and
% TODO: ref Document \ref{sec:supplement_rejected_pango_x_lineages}
% to mention multi breakpoints and close issue 379
Document \ref{sec:supplement_rejected_pango_x_lineages}
for an analysis of the Pango X lineages
which were detected as multiple-breakpoint recombinants by sc2ts
but were not added to the ARG
due to high HMM costs and/or poor sample support.

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG
(shading in Figure~\ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group
of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants
inferred within each week along the ARG,
focusing on the 354 events passing QC filters.
Recombination events were inferred more often
during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombination events rose with the global case count
(Figure~\ref{fig:recombinants_over_time}B; 
% JK is this correct?? we had 1.0 + 2.7 \; 10^{-7} x$ which I found confusing
linear regression: $1.0 + 2.7\times 10^{-7} x$, $p = 1.1 \times10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure~\ref{fig:recombinants_over_time}C; linear regression: 
$0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

@@

% P1. The ARG. Unique resource. Emphasise utility of UShER and how 
% this has more.
% SHZ: This makes it sound like we are just making analyses
% (which people can do already using existing methods and tools)
% easier to do
% rather than making a leap forward from anything else out there.
We present the first pandemic-scale ancestral recombination graph (ARG) for
SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
ARG represents a step change in how the evolutionary history of this virus can
be represented and analysed. Like the UShER phylogeny, it provides a
comprehensive and continuously updatable record of global genomic diversity,
but it extends this paradigm by embedding recombination directly into the
genealogy. The ARG unifies point mutations, deletions, and recombination events
within a single, compact data structure, enabling analyses that
were previously fragmented across multiple methods and data models. By encoding
reticulate ancestry explicitly rather than assuming a single tree, the ARG
provides a complete and internally consistent representation of SARS-CoV-2
evolution. In this way, the
ARG offers the same accessibility and scalability that made the UShER tree
indispensable during the pandemic, while expanding its scope to encompass the
full spectrum of evolutionary processes shaping SARS-CoV-2.

% P2. This is a conservative lower-bound on recombination. 
% Discuss other estimates.
% 
% SHZ: Why should we care about getting this number/frequency right?
% Sure, popgen folks love estimating this, but to other people,
% existing estimates already tell us that
% detectable recombination in SARS-CoV-2 is (qualitatively) rare,
% even without the advances of sc2ts.
% Also, we are talking about the frequency of detectable recombination
% in the last sentence, right?
% The true frequencey of recombination is unknowable from this type of data.
% So, the best we are ever going to get is a better lower bound.
The recombination frequency estimated from the \texttt{sc2ts} ARG is
substantially lower than that implied by the combination of UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
reported 589 unique recombination events with 43,104 descendant
samples---corresponding to roughly 2.7\% of all genomes being of recombinant
origin.  Using the same time interval, our inference at \(k=4\) identifies only
14 high-confidence recombination events with 172 descendants among 501,235
genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
explained by the stricter parsimony penalty adopted here: increasing the
penalty from \(p=3\) (as used in the RIPPLES analysis) 
to \(k=4\) greatly reduces false
positives while retaining nearly all verified recombinant lineages, 
including those missed by RIPPLES at equivalent thresholds.  
We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
of recombination as
tree rearrangements proceed, thereby requiring the more lenient 
threshold in order to find known recombinats.
These comparisons suggest that our estimate represents a conservative 
lower bound. Further work will be required to determine
the true frequency of recombination in SARS-CoV-2 evolution.

% P3. Clarifies relationships between pango X lineages, simplifying
% the Omicron sublineage.
% 
% SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
% Sure, phylogeneticists would get a kick out of it,
% but I don't think the majority of Cell readership would.
% I think what is important is knowing
% how many emerging recombinants carrying new combinations of mutations of concern
% we are actually dealing with.
% Clarifying the relationships between Pango Xs helps with this task.

@@

% P5. Confirm breakpoints around Spike. 
% SHZ: Maybe we can say it's a known (instead of robust)
% pattern of detectable recombination in CoVs in general?
% There are some citations for template switching hotspots in SC2.
% Also, it's not only about selection favoring new combinations of mutations in Spike.
% It's about combinations of mutations in, say, ORF1ab, getting selected against
% (or just low mutation rates).
% Maybe mention epistasis?
[FIXME: This is directish output from ChatGPT. Review!]
The genomic distribution of breakpoints in the ARG is highly uneven, with a
pronounced enrichment near the $5'$ boundary of the Spike gene and across the
S1–S2 junction.  This pattern closely mirrors that observed in the
UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
analyses \citep{Jackson2021} [more cites], but here it is recovered
independently over a much longer sampling period and under a more conservative
inference regime.  Under our more conservative estimate of the overall
prevalence of recombination, we find that this enrichment persists despite a substantial
reduction in the total number of inferred recombination events, indicating that
it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
by-product of detection sensitivity.  Across all high-confidence events, the
median breakpoint interval length is $\sim 2.2$kb and declines with parental
divergence, consistent with expectations from sequence-based localization
accuracy.  These results show that even with a conservative lower bound on
recombination frequency, the concentration of breakpoints around Spike remains
a robust signature of coronavirus genome biology—likely reflecting a
combination of mechanistic biases in template switching and selective
constraints on Spike function.

@@

%%% WIP: SHZ
We present the first pandemic-scale
ancestral recombination graph (ARG) for SARS-CoV-2,
reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
We demonstrate how this ARG, for the first time,
enables systematic identification and characterization of
recombination events in fine-grained
detail--in terms of their parents, breakpoints, and timing of
emergence--providing richer and more nuanced
genealogical, genomic, and epidemiological context
than existing phylogenetic approaches.
Together, sc2ts and the ARG represent a new paradigm in
how we study the evolution of pathogens in the face of ongoing sampling.

Accounting for recombination is a key challenge in pathogen genomics, and
recombination has been largely ignored in analysis of viral genomes.
Existing phylogenetic approaches for pathogens detect recombination
\textit{post hoc} (e.g., UShER+RIPPLES),
treat it as a nuance (e.g., XX), or
discard its signals from downstream analyses (e.g., XX).
Sc2ts addresses this challenge
by using the ARG as the underlying data model,
explicitly representing recombination events as individual nodes and
relating them to specific parents in the ARG.
The ARG provides a coherent framework to determine and represent
the relationships between recombinants and their specific parents
(which are not possible using a tree-based data model,
e.g., the Mutation Annotated Tree format of UShER).
The ARG reveals and clarifies the relationships between
individual samples of the same recombinant origin,
avoiding issues such as over-naming recombinants
(e.g., the samples assigned to
five differently labeled recombinant Pango lineages
descending from the “Xx” recombination event in the ARG).
Our results show how an ARG can considerably simplify
efforts to identify and monitor emerging recombinants,
facilitating communication of genomic surveillance results.

% SHZ: A bit vague, but is this worth discussing?
% What else to say about the timing of these events?
% TODO: Example with say XBB?
% TODO: Say we don’t explore this fully, but it could be a follow-up study?
The timing of recombination events is of key interest in
pathogen genomic epidemiology,
because this information may yield insights into
% Probably A bad idea to say this.
how the emergence of a new recombinant might be
linked with changes in public health policies.
Using the ARG, we reconstructed a historical timeline of
recombination events during the COVID-19 pandemic,
showing that it is possible to pinpoint the occurrence of
recombination events precisely in time,
within windows of X to X months
(given the intensity of sampling in the COVID-19 dataset).

The overall frequency of detectable recombination events
estimated from the sc2ts ARG is substantially lower than
that from UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
When applied \textit{post hoc} to an UShER phylogeny of
$\sim$1.6 million genomes sampled up to May, 2021,
RIPPLES reported 589 unique recombination events with 43,104 descendant
samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
Using the same time interval, sc2ts \(k=4\) identifies
14 high-confidence recombination events with
172 descendant samples among 501,235 samples \(0.034\%\),
a 40- to 80-fold reduction.
Much of this difference is explained by
the stricter parsimony penalty used here:
increasing the penalty from p=3 (as used in the RIPPLES analysis) to
k=4 greatly reduces false positives
while retaining nearly all verified recombinant lineages,
including those missed by RIPPLES at equivalent thresholds.
These indirect comparisons suggest that
our estimate represents a conservative lower bound.
In future studies, dissecting the methodological differences
between sc2ts and UShER+RIPPLES may
inspire new ideas to develop or improve ARG building methods
for SARS-CoV-2 (and potentially other pathogens).

The ARG itself is a powerful resource
to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies
important features of SARS-CoV-2 evolution,
which are shown in different studies
using various methods, tools, and approaches.
The ARG also compactly encodes genetic variation in
millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of
this valuable dataset for a range of
genetic and evolutionary analyses beyond those presented here,
for instance, to understand
how the variability in sequencing efforts
across time and location affects our ability
to reconstruct the origin of important recombinant lineages and
to estimate when they emerge.
Because building the ARG is computationally expensive, and
checking and improving various aspects of its quality is laborious,
we hope that making this resource openly available will
save significant time and effort for other researchers,
enabling immediate focus on analyzing the information
contained within the COVID-19 dataset.
% TODO: Briefly discussion software tools here or in a subsequent paragraph.

Recent developments in ARG methods have focused almost exclusively on
applications to human sequencing data.
The availability of human ARGs reconstructed from large datasets
has already had a tremendous impact on improving
our understanding of human genetic ancestry and past demography.
We anticipate similar advances to be made
by adapting existing inference methods for human ARGs
to use the sc2ts ARG for the genealogy-based study of
natural selection or the geographic spread of SARS-CoV-2 through time.

The unprecedented throughput and pace of genome sequencing are
creating ample opportunity to understand the impacts of recombination
on pathogen evolution and epidemiology.
We hope that this study will spur
development of ARG methods for pathogens,
ultimately making recombination-aware genome analysis
routine for all sorts of pathogens beyond SARS-CoV-2.

@@

We close by emphasizing several current limitations
as well as potential directions to improve sc2ts,
which we hope will also inspire future research
to develop ARG methods for other pathogens.

Deletions which are reverted can be strong signals of recombination,
because bases which are lost cannot be easily regained,
except through recombination involving
a lineage that has bases at the deleted sites.
We currently treat multi-base deletions as missing data.
More sophisticated treatments of such indels in the HMM
may help recover additional signals
(and avoid false signals) of recombination.

The current HMM assumes a simplistic mutation model
(where mutation to any other allele is equally probable) and
implicitly assumes a uniform mutation rate along the SARS-CoV-2 genome.
However, genome analyses have revealed
transition-to-transversion rate bias in SARS-CoV-2 \cite{Bloom2023} and
sequence context-dependent variation
in the mutation rate along the SARS-CoV-2 genome (REF).
A more realistic mutation model may help improve
inference of local genetic relationships in the ARG, and
accounting for mutation rate heterogeneity
during sample matching or in the HMM cost
(e.g., down-weighing sites in highly mutable regions)
may help reduce false positive recombinants.

We developed and used several methods
to exclude poor-quality recombination events post hoc, including
filtering by supporting loci per parent and
long-branch splitting.
These methods could instead be run immediately
after sample matching during ARG inference,
obviating the need for post-processing of the ARG.

An inherent limitation of detecting recombination from sequence data,
which applies to all methods,
is that it is impossible to detect events with few (or no) supporting bases,
as these cannot be distinguished from mutation events.
This limitation makes it particularly challenging
to detect events near the termini of linear viral genomes.

It is challenging to accurately reconstruct global events
when there are regions with poor sampling and jurisdictions
with long delays before sequences become publicly available (REF).
Revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating
with samples that became available later) and
better handling of putative recombinants
with suspiciously long branches (e.g., Delta and Omicron BA.2)
might improve the accuracy of the genealogy.

% TODO: Add a short discussion of the node dates and
% some references to the literature, closing issue 725.

Finally, future developments may account for geography in the HMM,
favoring attachment of samples to geographically close lineages.

@@

We validated recombination events in the sc2ts ARG 
using 3SEQ~\cite{Lam2018improved} as follows.
We first exported the alignments for the recombination node and both 
parents to FASTA format, and then ran 3SEQ in ``full mode''
independently for each recombinant. Specifically, we 
ran \texttt{3seq -full parents.fa recombinant.fa}
% TODO: Should we mention the default p-value threshold is 0.05?
using default thresholds.
We used 3SEQ source code downloaded from \url{https://gitlab.com/lamhm/3seq}
at git hash e45ee3b4.

@@

Recombination breakpoint intervals are imprecise
when the recombining parents are identical in sequence
at sites spanning the breakpoint.
Here, we estimate the interval lengths expected around a breakpoint,
given a genome-wide nucleotide substitution rate ($\mu$),
total time between two samples ($T$), and
sufficient support to detect the recombination event.
If a breakpoint occurs a fraction $P$
along a genome of length $L$ and
$k$ mutations have occurred to the left of the breakpoint,
then we expect the nearest mutation
to occur $P L / (k + 1)$ bases to the left.
If we require $n$ mutations on both sides of the breakpoint,
we can calculate the expected value of the interval to the left
by summing across a Poisson-distributed number of mutations,
conditioned on having at least $n$ mutations,
using the following equation
(where $\Gamma(x)$ is the gamma function, and
$\Gamma(x, y)$ is the upper incomplete gamma function):

@@

To explore the timing of emergence of recombination events,
case count data were downloaded from
the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE)
at Johns Hopkins University \cite{Dong2020} and
binned by week from January, 2020 to March, 2023.
The proportion of each lineage $i$ in each week ($p_i$) was then inferred
from the Scorpio designation at each node within the ARG for that week.
Genetic variation was measured as the ``expected heterozygosity'' across lineages,
which is the chance that two lineages chosen at random
from that week had different Scorpio designations ($\sum_{i} \sum_{j>i} 2 p_i p_j$).
The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time}
allowed for a non-zero intercept,
because there may be unreported cases even without any reported cases.
The slopes were highly significant, and
the p-values were confirmed by randomly permuting the predictor variable 10,000 times.
ANOVAs were conducted to determine which single predictor
($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
These statistical tests and graphics were conducted
% TODO: The link!
in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.

In addition to the analysis of the 354 recombination events that pass QC,
we analyzed the relationship between the full set of 855 recombinants and case numbers.
These additional recombinants occurred disproportionately
during the Delta and first Omicron wave (July, 2021 to February, 2022).
Again, the number of recombinant events rose with the global case count
(linear regression: $3.1 + 5.3 \; 10^{-7} x$, $p = 0.0006$)
and with the diversity of lineages in each week
(linear regression: $0.42 + 18.3 x$, $p = 2 \; 10^{-12}$).
The single best predictor of recombination events
among the predictors tested ($n$, $H$, $n \; H$, $n^2$, $n^2 H$)
was again the product of cases and diversity ($n \; H$),
with an adjusted $R^2$ of 26.9\%,
although this was now followed closely by expected heterozygosity on its own ($H$),
with an adjusted $R^2$ of 26.5\%.
Thus, similar conclusions are reached with either dataset,
although filtering the recombination events to the 354
with better support revealed a stronger relationship with case numbers.

@@

\subsection{Pango X lineages not added to the ARG}
\label{sec:supplement_rejected_pango_x_lineages}
